Equalizer Programming Guide

### 2008-05-02 Developer/ProgrammingGuide/paper.tex
Website: improve statistics docs, add statistics chapter to Programming Guide, improve PG images based on print version
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.2 for Equalizer 0.5\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2em}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
  \vspace{-2em}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\subsection{Event Handling}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=3cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=3cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=3cm]{images/tracked}}%
  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
    frustra}}
\end{figure}

@@

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.1 for Equalizer 0.5+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1.5em}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
  \vspace{-1.5em}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\subsection{\label{sEventHandling}Event Handling}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=3.5cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=3.5cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=3.5cm]{images/tracked}}%
  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
    frustra}}
\end{figure}

@@

\subsection{Statistics}

\subsubsection{Statistics Gathering}

Statistics are measured in milliseconds since the configuration was
initialized. On each node, the global configuration clock is reset at
the same time by the server during \textsf{Config::init}. Each statistic
event records the originator's (channel, window or config) unique
identifier.

Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
The events are processed by the channel's and window's
\textsf{processEvent} method. The default implementation sends these
events to the config, using \textsf{Config::sendEvent}, as explained in
\sref{sEventHandling}. When the default implementation of
\textsf{Config::handleEvent} receives the statistics event, it sorts the
event per frame and per originator. When a frame has been finished, the
events are pushed to the local (app-)node for visualization.

\fig{fStatistics} shows the visualization of statistics events in an
overlay\footnote{3D model courtesy Stanford University Computer Graphics
  Laboratory.}.

\subsubsection{Statistics Overlay}

In debug builds, a statistics overlay using the gathered statistics
events is rendered directly before \textsf{Window::swapBuffers}. The
overlay rendering method, \textsf{Channel::drawStatistics}, can be
called directly to render the overlay. It can also be overwritten by a
custom implementation.

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
    {\caption{\small\label{fStatisticsDetail}Annotated Detail of the
        Statistics from \fig{fStatistics}.}}
\end{figure}

\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics},
together with some annotations for the overlay. The statistics shown are
for a two-node 2D compound. The destination channel is on the
\textsf{appNode} and contributes to the rendering.

This compound executes two \textsf{Channel::frameDraw} tasks, one
\textsf{Channel::frame\-Read\-back} task on the remote node and one
\textsf{Channel::frameAssemble} task on the local node.

The X axis is the time. One pixel on the screen corresponds to one
millisecond. The right-most pixel is the most current time. On the Y
axis are the rendering entities: channels, windows and the config. The
upper channel is the local channel since it executes
\textsf{frameAssemble}, and the lower channel is the remote channel,
executing {frameReadback}.

The remote machine is a Mac Mini, the local a MacBook Pro. The draw
times are unbalanced due to the speed difference of the machines, and
the transmit time is relatively long since 100 MBit Ethernet was used.

In order to facilitate the understanding, subsequent frames alternate in
brightness. The right-most, current frame is brighter than the
frame-before-last.

The configuration used has a latency of one frame. Consequently, the
execution of two frames overlaps. This can be observed in the early
execution of the remote channel's \textsf{frameDraw}, which starts while
the local channel is still assembling the previous frame.

The beginning of a frame is marked by a vertical green line, and the end
of a frame by a vertical gray line. These lines also alternate in
brightness. The brightness and color matches the event for
\textsf{Config::startFrame} and \textsf{Config::finishFrame},
respectively. The event for \textsf{startFrame} is typically not
visible, since it takes less than one millisecond to execute. If no idle
processing is done by the application, the event for
\textsf{finishFrame} occupies a full frame, since the config is blocked
here waiting for the frame to complete.

In the above example, the local channel finishes drawing the frame
early, and therefore spends a considerable amount of time waiting for
the input frame from the remote channel. These wait events, rendered
red, are a sub-event of the yellow \textsf{frameAssemble} task.

The white \textsf{Window::swapBuffers} task takes also a long time in
debug builds, since the execution of the swap buffer is locked to the
vertical retrace of the display, and enforced by \textsf{glFinish}
before sampling the end time of the event. In release builds, this does
not happen and all OpenGL commands are pipelined with the execution on
the CPU.
